---
title: |
  | &nbsp;
  | Importing, working with, and exploring data
  | Week 2, Lecture 03:
author: |
  | &nbsp;
  | Richard E.W. Berl
date: |
  | Spring 2019
  | &nbsp;
output:
  html_document:
    highlight: haddock
    theme: flatly
  pdf_document: 
    highlight: haddock
---

## Loading data

### From a package

```{r}
data("HairEyeColor")
```

```{r, eval=FALSE}
?HairEyeColor
```

```{r}
HairEyeColor
```

```{r}
hairEyeColor = as.data.frame(HairEyeColor)
hairEyeColor
rm(HairEyeColor)
```

### From a CSV

### From a tab-delimited file

Visit: [http://www.randomservices.org/random/data/HorseKicks.html](http://www.randomservices.org/random/data/HorseKicks.html)

Read the description of the data set.

Click the highlighted text "Horse-kick data" at the bottom of the page to download it. Place it in your class `/data` directory (or wherever you are placing your raw data for the course).

```{r}
getwd()
# horseKicks = read.table(file="./data/HorseKicks.txt", header=TRUE, sep="\t")
```



## Manipulating data

### Subsetting

Subsetting is one of the most common tasks when you're dealing with data. There are multiple ways to subset data, specifically when it is in a data frame.

## Cleaning data

### Finding and fixing missing data

## Problems with data

### Missing data

### Confounds



## What is EDA?

### EDA is NOT "fishing," "data dredging," or "p-hacking"


## Descriptive statistics

### Confidence intervals