---
title: |
  | &nbsp;
  | Week 6, Lecture 11
  | Advanced statistical methods, part II: Structural equation modeling, social network analysis, and geospatial analyses
author: |
  | &nbsp;
  | Richard E.W. Berl
date: |
  | Spring 2019
  | &nbsp;
output:
  html_document:
    highlight: haddock
    theme: flatly
  pdf_document:
    highlight: haddock
---

## Structural equation modeling

Bob Marshall Wilderness data: https://www.fs.usda.gov/rds/archive/Product/RDS-2017-0016

Instructions from before:
At the bottom, click "Download data publication," which gives you a ZIP archive. Open it up, go into the "Data" folder and pull out both CSVs for your `/data` directory. You can hang on to the other files in the archive as well, for the metadata.

This time, load in the mailback survey data:

```{r}
mail = read.csv("./data/BMWC2004_maildata.csv", header=T, na.strings=c("88","99"),
                stringsAsFactors=F)
```

```{r}
mail[1:5,1:10]
str(mail)
colnames(mail)
```

```{r}
mail$q11feel
mail$q11feel[mail$q11feel == 6] = NA
mail$q11feel = ordered(mail$q11feel)

mail[,c(105:136,139:174)] = lapply(mail[,c(105:136,139:174)], function(x) ordered(x))
```

<br>

### Exploratory factor analysis

```{r, warning=FALSE}
library(psych)
```

```{r, warning=FALSE}
library(lavaan)
```

```{r, eval=FALSE}
install.packages("semTools")
```

```{r, warning=FALSE}
library(semTools)
```

```{r}
set.seed(20190430)
trainSet = sample(nrow(mail), round(nrow(mail)/2))
mailTrain = mail[trainSet,]
mailTest = mail[-trainSet,]
```


```{r, error=TRUE}
mailCor = lavCor(mailTrain[,c(48,105:136,139:174)])
```

```{r, error=TRUE}
mailCor = lavCor(mailTrain[,c(105:121,123:136,139:174)])
```

```{r}
mailCor = lavCor(mailTrain[,c(105:121,123:126,128:136,141:155,
                              157,158,161:164,166:173)])
dim(mailCor)
mailCor[1:10,1:10]
```

```{r, eval=FALSE}
?semTools::efaUnrotate
```

```{r}
fa.parallel(mailCor, fm="wls", n.obs=nrow(mailTrain))
```

```{r}
mailEFA = efaUnrotate(mailTrain[,c(105:121,123:126,128:136,141:155,
                                   157,158,161:164,166:173)],
                      nf=14, estimator="WLSMV", start=F)
```

<br>

### Confirmatory factor analysis




<br><br>


([pdf](./lecture10.pdf) / [Rmd](./lecture10.Rmd))

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>