---
title: |
  | &nbsp;
  | Week 6, Lecture 12
  | Advanced statistical methods, part II: Structural equation modeling, social network analysis, and geospatial analyses
author: |
  | &nbsp;
  | Richard E.W. Berl
date: |
  | Spring 2019
  | &nbsp;
output:
  html_document:
    highlight: haddock
    theme: flatly
  pdf_document:
    highlight: haddock
---

## Social network analysis

Download Ancestral Pueblo food web network: https://www.sciencedirect.com/science/article/pii/S0305440317300377#appsec1

"Download spreadsheet (329KB)"

```{r, warning=FALSE}
library(readxl)
library(igraph)
```

```{r}
foodN = as.data.frame(read_xlsx("./data/1-s2.0-S0305440317300377-mmc1.xlsx", sheet=1))
foodE.AP = as.data.frame(read_xlsx("./data/1-s2.0-S0305440317300377-mmc1.xlsx", sheet=2,
                                   range="G2:H1398"))
```

```{r}
head(foodN)
str(foodN)

head(foodE.AP)
str(foodE.AP)
```

```{r}
foodN$Class = substr(foodN$Species, 1, 1)
foodN$Species = substr(foodN$Species, 3, nchar(foodN$Species))
```

```{r}
foodN.AP = foodN[foodN$ID %in% unique(unname(unlist(foodE.AP))),]
```

```{r}
foodE.AP = foodE.AP[,c(2,1)]
```

```{r}
foodNet = graph_from_data_frame(foodE.AP, vertices=foodN.AP)
foodNet
```

```{r, fig.width=8, fig.height=8}
plot(foodNet)
```

```{r, eval=FALSE}
?igraph::igraph.plotting
```

```{r, fig.width=8, fig.height=8}
plot(foodNet,
     vertex.size=6, vertex.shape="none",
     vertex.label.cex=ifelse(V(foodNet)$Species == "human", 1, 0.75),
     vertex.label.font=ifelse(V(foodNet)$Species == "human", 2, 1),
     vertex.label.color=as.numeric(as.factor(V(foodNet)$Class)),
     edge.arrow.size=0.5, edge.color=adjustcolor("darkgrey", alpha.f=0.15),
     layout=layout_with_kk)
```

<br>

### Descriptive statistics

**Edge density**

Ratio of the number of edges to all possible edges

```{r}
edge_density(foodNet, loops=F)
```

**Reciprocity**

Proportion of mutual connections

```{r}
reciprocity(foodNet, ignore.loops=T)
dyad_census(foodNet)
```

**Transitivity**

Probability of three vertices being connected

```{r}
transitivity(foodNet, type="global")
triad_census(foodNet)
```

```{r, eval=FALSE}
?triad_census
```

**Diameter and distance**

Longest chain of connections

```{r}
diameter(foodNet, directed=T)
get_diameter(foodNet, directed=T)
```

Bottle gourd -> human -> dog -> tick -> sage grouse -> coyote -> snowshoe hare

```{r}
mean_distance(foodNet, directed=T)
```

Shortest distance from humans to pinyon jay

```{r}
shortest_paths(foodNet, from="96", to="43", output="both")
```

Human -> dog -> tick -> centipede -> pinyon jay

<br>

#### Centrality

**Degree**

Number of connections

```{r}
degree(foodNet, v="96", mode="out")
degree(foodNet, v="96", mode="in")
```

```{r}
ego(foodNet, nodes="96", mode="out")
```

Domestic dog

```{r}
ego(foodNet, nodes="96", mode="in")
```

**Closeness**

Inverse distance to all other nodes

```{r}
closeness(foodNet, vids="96", mode="all")
closeness(foodNet, mode="all")[which.max(closeness(foodNet, mode="all"))]
```

Colorado chipmunk

**Betweenness**

Number of paths passing through a node

```{r}
betweenness(foodNet, v="96", directed=T)
betweenness(foodNet, directed=T)[which.max(betweenness(foodNet, directed=T))]
```

**Eigenvector**

Values of the first eigenvector

```{r}
eigen_centrality(foodNet, directed=T)$vector[names(eigen_centrality(foodNet, directed=T)$vector) == "96"]
eigen_centrality(foodNet, directed=T)$vector[which.max(eigen_centrality(foodNet,
                                                                        directed=T)$vector)]
```

Domestic dog

**Hub**

```{r}
hub_score(foodNet)$vector[names(hub_score(foodNet)$vector) == "96"]
hub_score(foodNet)$vector[which.max(hub_score(foodNet)$vector)]
```

Maize

**Authority**

```{r}
authority_score(foodNet)$vector[names(authority_score(foodNet)$vector) == "96"]
authority_score(foodNet)$vector[which.max(authority_score(foodNet)$vector)]
```

Mule deer

<br>

#### Community detection

```{r, eval=FALSE}
?igraph::communities
```

```{r}
foodNetU = as.undirected(foodNet, mode="mutual")
```

**By greedy optimization of modularity**

```{r}
foodNetCFG = cluster_fast_greedy(foodNetU)
foodNetCFG
dendPlot(foodNetCFG, mode="hclust")
```

**By edge betweenness**

```{r}
foodNetCB = cluster_edge_betweenness(foodNetU)
foodNetCB
dendPlot(foodNetCB, mode="hclust")
foodNetCB[20]
```




([pdf](./lecture12.pdf) / [Rmd](./lecture12.Rmd))

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>