---
title: |
  | &nbsp;
  | Importing, working with, and exploring data
  | Week 2, Lecture 03
author: |
  | &nbsp;
  | Richard E.W. Berl
date: |
  | Spring 2019
  | &nbsp;
output:
  html_document:
    highlight: haddock
    theme: flatly
    toc: yes
    toc_float: yes
  pdf_document:
    highlight: haddock
    toc: yes
---

## Loading data

### From a package

For a full list of data sets included in the base `datasets` R package, use: `data()` or `library(help="datasets")`.

You can load a built-in data set like this:

```{r}
data("HairEyeColor")
```

There is documentation available for each one.

```{r, eval=FALSE}
?HairEyeColor
```

Now let's examine the data.

```{r}
HairEyeColor
```

Whoa, this isn't a data frame. What is it?

```{r}
class(HairEyeColor)
```

It's a three-way contingency table. This makes it easier to look at and is suitable for some analyses. But we can coerce it to a data frame, which will make it easier for us to work with:

```{r}
hairEyeColor = as.data.frame(HairEyeColor)
hairEyeColor
```

Much better. And since we renamed the data frame (to the "mixedCase" style), we can remove the old object:

```{r}
rm(HairEyeColor)
```


### From a CSV file

Visit: [https://osf.io/s7d9d/](https://osf.io/s7d9d/)

Read the description of the data set. The source of the data is:  

* Weiss, A., et al. (2017). Personality in the chimpanzees of Gombe National Park. Scientific Data, 4, 170146. doi: [10.1038/sdata.2017.146](https://doi.org/10.1038/sdata.2017.146)

You can also click "Codebook for Gombe data personality variables.pdf" in the "Files" box of the OSF page to learn about how the data are coded.

In the "Files" box, click "gombe_128.csv" and then the "Download" button on the top right of the following page to download it. Place it in your class `/data` directory (or wherever you are placing your raw data for the course).

You can open the file (in Excel or a text editor) to see how it is formatted.

```{r}
gombe = read.csv(file="./data/gombe_128.csv", header=TRUE)
```

```{r}
head(gombe)
```

### From a tab-delimited file

Visit: [http://www.randomservices.org/random/data/HorseKicks.html](http://www.randomservices.org/random/data/HorseKicks.html)

Read the description of the data set.

At the bottom of the page, click the highlighted text "Horse-kick data" to download it. Place it in your class `/data` directory (or wherever you are placing your raw data for the course).

You can open the file to see how it is formatted.

```{r}
horseKicks = read.table(file="./data/HorseKicks.txt", header=TRUE, sep="\t")
```

```{r}
horseKicks
```

### From an Excel spreadsheet

Visit: [https://royalsocietypublishing.org/doi/suppl/10.1098/rsos.150645](https://royalsocietypublishing.org/doi/suppl/10.1098/rsos.150645)

You can click "View Full Text" on the left side to read the article (or just the abstract) to learn about the data set.

Under the "Supplemental Material" heading, click the highlighted text "rsos150645supp1.xlsx" to download it. Place it in your class `/data` directory.

You can open the file to see how it is formatted.

```{r, eval=FALSE}
install.packages("tidyverse")
# OR
install.packages("readxl")
```

```{r}
library(readxl)
```

Documentation is available at: [https://readxl.tidyverse.org/](https://readxl.tidyverse.org/)

```{r}
folktales = read_xlsx(path="./data/rsos150645supp1.xlsx",
                      sheet=1, range="A2:JP52")
folktales
```

```{r}
folktales = as.data.frame(folktales)
folktales[1:5,1:15]
```

```{r}
colnames(folktales)[1] = "society"
folktales[1:5,1:15]
folktales$society
```


<br><br>


([pdf](./lecture03.pdf) / [Rmd](./lecture03.Rmd))

<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>